Is it time for an NBA expansion?

### 2024-04-21 paper/paper.qmd
Model discussion

library(tidyverse)
library(knitr)
library(modelsummary)
library(tinytex)
library(lubridate)
library(viridis)
```

@@

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

current_nba_data_long <- nba_data %>%
  filter(Year >= 2004) %>%  # Keep only years from 2004 onwards
  pivot_longer(cols = c(AST, TRB, PTS, STL, BLK, TOV), names_to = "Statistic", values_to = "Value")

current_plot <- ggplot(current_nba_data_long, aes(x = Year, y = Value)) +
  geom_point(color = "black") +  # All points in black
  geom_smooth(method = "lm", aes(color = Statistic), se = FALSE) +  # Lines of best fit with different colors
  scale_color_manual(values = line_colors) +
  facet_wrap(~Statistic, scales = "free_y") +  # Facet by statistic, with free y scales
  labs(title = "NBA Stats Over Years (Post-2004)", x = "Year", y = "Value") +
  theme_minimal() +
  theme(legend.position = "none")

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

ggplot(scoring_data, aes(x = Year, fill = factor(Year))) +
  geom_histogram(stat = "count", binwidth = 1, color = "black") +
  scale_fill_viridis(discrete = TRUE, option = "mako") +  # Using 'magma' for a warm color palette
  labs(title = "Frequency of Records by Year",
       x = "Year",
       y = "Count of Records",
       fill = "Year") +
  theme_minimal() +
  theme(legend.position = "none")

@@

```{r}

@@

ggplot(data_1980_onwards, aes(x = Year, fill = as.factor(Year))) +
  geom_histogram(stat = "count", bins = length(unique(data_1980_onwards$Year)), color = "black") +
  scale_fill_viridis(discrete = TRUE) +  # Apply viridis color scale for discrete data
  labs(title = "Count of Records per Year from 1980 Onwards",
       x = "Year Modified",
       y = "Count of Records") +
  theme_minimal() +
  theme(legend.position = "none")

@@

```{r}
##
data_1980_70pts_plus <- scoring_data %>%
  filter(Year >= 1980, PTS >= 70)

ggplot(data_1980_70pts_plus, aes(x = Year, fill = factor(Year))) +
  geom_histogram(stat = "count", bins = length(unique(data_1980_70pts_plus$Year)), color = "black") +
  scale_fill_viridis_d(begin = 0.2, end = 0.8, direction = 1, option = "magma") +  # Discrete viridis scale
  labs(title = "Count of Records per Year for Scores of 70+ PTS from 1980 Onwards",
       x = "Year Modified",
       y = "Count of Records") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

ggplot(comparison_data, aes(x = Year, y = Ratio)) +
  geom_point(aes(color = Year), size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "navy") +
  labs(title = "LeBron's PTS as a Ratio to NBA Average PTS Across Seasons",
       x = "Season",
       y = "Ratio of LeBron's PTS to NBA Average PTS") +
  theme_minimal()

ggplot(comparison_data, aes(x = Year, y = Ratio, fill = Year)) +
  geom_bar(stat = "identity", color = "black") +
  labs(title = "LeBron's PTS as a Ratio to NBA Average PTS Across Seasons (Bar Graph)",
       x = "Season",
       y = "Ratio of LeBron's PTS to NBA Average PTS") +
  theme_minimal() +
  theme(legend.position = "none")

@@

@@

msummary(
  list("Points Model" = nba_model),
  fmt = 2
  )
```

>>>>

library(tidyverse)
library(knitr)
library(modelsummary)
library(tinytex)
library(lubridate)
library(viridis)
library(kableExtra)
```

@@

The goal of our modelling strategy is twofold. Firstly, I want to analyze what how much other basic statistics such as assists and rebounds have on PPG as well as the affect the number of teams in the league in comparison with the year have on the same points per game. Next, I would like to predict the PPG of the next three seasons depending on the number of teams that are in the league. Considering the NBA would likely expand only one team at a time we will be focusing on the potential PPG over the next three years with 30 teams, 31 teams, and 32 teams.

Here we briefly describe the Bayesian analysis model used to investigate the multiple linear regression model of PPG as described below.

@@

current_nba_data_long <- nba_data %>%
  filter(Year >= 2004) %>%  
  pivot_longer(cols = c(AST, TRB, PTS, STL, BLK, TOV), names_to = "Statistic", values_to = "Value")

current_plot <- ggplot(current_nba_data_long, aes(x = Year, y = Value)) +
  geom_point(color = "black") +  
  geom_line(aes(color = Statistic)) + 
  scale_color_manual(values = line_colors) +
  facet_wrap(~Statistic, scales = "free_y") +  
  labs(title = "NBA Stats Over Years (Post-2004)", x = "Year", y = "Quantity") +
  theme_minimal() +
  theme(legend.position = "bottom")

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-scoringgraph
#| fig-cap: High Scoring games(59+ pts)

ggplot(scoring_data, aes(x = Year, fill = factor(Year))) +
  geom_histogram(stat = "count", binwidth = 1, color = "black") +
  scale_fill_viridis(discrete = TRUE, option = "mako") +  # Using 'magma' for a warm color palette
  labs(title = "Count of 59+ Point Games",
       x = "Year",
       y = "Count of Games",
       fill = "Year") +
  theme_minimal() +
  theme(legend.position = "none")

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-scoringgraph1980
#| fig-cap: High Scoring games(59+ pts) Since 1980

@@

ggplot(data_1980_onwards, aes(x = Year, fill = as.factor(Year))) +
  geom_histogram(stat = "count", bins = length(unique(data_1980_onwards$Year)), color = "black") +
  scale_fill_viridis(discrete = TRUE) +  
  labs(title = "Count of 59+ Point Games since 1980",
       x = "Year Modified",
       y = "Count of Games") +
  theme_minimal() +
  theme(legend.position = "none")

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-scoringgraph70
#| fig-cap: High Scoring games(70+ pts) Since 1980

data_1980_70pts_plus <- scoring_data %>%
  filter(Year >= 1980, PTS >= 70)

ggplot(data_1980_70pts_plus, aes(x = Year, fill = factor(Year))) +
  geom_histogram(stat = "count", bins = length(unique(data_1980_70pts_plus$Year)), color = "black") +
  scale_fill_viridis_d(begin = 0.2, end = 0.8, direction = 1, option = "magma") +  # Discrete viridis scale
  labs(title = "Count of 70+ Point Games since 1980",
       x = "Year Modified",
       y = "Count of Games") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-lebronbar
#| fig-cap: Lebrons PPG as a Ratio of the League Average

@@

ggplot(comparison_data, aes(x = Year, y = Ratio, fill = Year)) +
  geom_bar(stat = "identity", color = "black") +
  labs(title = "LeBron's PTS as a Ratio to NBA Average PTS Across Seasons",
       x = "Year",
       y = "Ratio of LeBron's PTS to NBA Average PTS") +
  theme_minimal() +
  theme(legend.position = "none")

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-lebronarea
#| fig-cap: League Average PPG vs Lebrons PPG

# Calculate average points per season for the NBA
nba_avg_pts <- nba_data %>%
  group_by(Year) %>%
  summarize(Avg_PTS = mean(PTS, na.rm = TRUE))

fa_year <- c(2010, 2014, 2018)

# Prepare comparison data by joining LeBron's data with NBA average points
comparison_data <- lebron_data %>%
  rename(Year = Season) %>%
  left_join(nba_avg_pts, by = "Year")

# Create an area graph for LeBron's points and NBA average points
ggplot(comparison_data, aes(x = Year)) +
  geom_area(aes(y = Avg_PTS), fill = "blue",alpha = 0.5, color = "navy") +
  geom_area(aes(y = PTS), fill = "maroon", alpha = 0.8, color = "darkred" ) +
  labs(title = "Comparison of LeBron's PTS vs. NBA Average PTS Across Seasons",
       x = "Season",
       y = "Points") +
  geom_vline(xintercept = fa_year, color = "black", linetype = "dashed") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 12)) +
  scale_fill_manual(values = c("blue", "maroon"),
                    labels = c("NBA Avg PTS", "LeBron PTS"),
                    name = "Legend")
```

@@

modelsummary(
  list("Points Model" = nba_model),
  fmt = 2
  )
```
